---
title: "Project Proposal"
author: 'Alexis Hegarty, Ella Smith, Ashlyn Philips, and Amy Truong'
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r}
#| code-fold: true
library(tidyverse)
library(broom)

co2 <- read_csv("co2.csv", col_names = TRUE)
cdmt <- read_csv("cdmt.csv", col_names = TRUE)
```

# Part 1: Introduction

### Data Description:

The data we are using in our analysis is sourced from [Gapminder](https://www.gapminder.org/data/documentation/). The Gapminder data sets are compiled of world data ranging from 1800 to 2020. The variables, documented by country (using current geographic boundaries), are indicators of global development and evolution through the past two centuries. Examples of these variables include, child mortality, GDP, fertility rates, life expectancy rates, and more.

The data sets that we will be using in our analysis include two quantitative variables: Carbon Dioxide (CO2) and Child Mortality (CDMT).

The Child Mortality Rate data set combines data from multiple sources to create a data set of child deaths under the age of 5 for each country from 1800 to 2100. Child mortality is measured as the number of 0-5 years dying per 1000 children born. The data set has three variables: country, year, and child mortality rate.

The Carbon Dioxide data set combines data from multiple sources to create a data set of CO2 emissions in tonnes per capita for each country from 1800 to 2022. The data set has three variables: country, year, and CO2 emissions per capita. Both data sets we are using contain from 194 countries.

### Hypothesis

We will use CO2 as our explantory variable and Child mortality as our response varaible.

We think that there maybe a relationship between these two variables. We understand that in general CO2 emissions are harmful to humans. From 1800 to 2022 as the world became more industrialized and CO2 emissions per capita exponentially, this also increased child mortality.

However, we are also aware that industrialization also could have lowered child mortality rates through the increase in medicine and technology.

Our investigation aims to explore this nuanced relationship between CO2 emissions and child mortality. Using historical data and statistical analyses, we aim to provide a more comprehensive understanding of the evolution of the relationship between these two variables within the last two decades.

### Clean & Merge Data

A glaring factor within the data is the child mortality dataset contains data all the way to 2100. This means the data past 2018 (noted in data description in Gapminder) is predicted. For the purposes of this analysis and the data we have for CO2 emissions, we will only use data from 1800 to 2018.

In our data cleaning process we will omit observations with at least one variable missing. We are aware that omitting these missing variables decreases the sample size and may lead us to miss important information. Since we are using ggplot and lm--which drop missing observations anyway--it makes sense to remove these variables for the entire analysis.

The two data sets will also need to be merged into one data set in long format using the pivot_longer function.

Another important factor to address in our data cleaning process is the different scales the two variables are measured by. "Deaths per 1000 children" is hard to compare in context to "tonnes per capita". To contextualize the two scales in comparison to each other we will use a log base transformation.



```{r}
#| code-fold: true

co2_cdmt <- cdmt %>%
  mutate(across(-country, as.character)) %>%
  pivot_longer(cols = -country, names_to = 'year', values_to = 'cdmt') %>%
  mutate(year = as.character(year)) %>%
  
  full_join(
    co2 %>%
      mutate(across(-country, as.character)) %>%
      pivot_longer(cols = -country, names_to = 'year', values_to = 'co2') %>%
      mutate(year = as.character(year)),
    by = c("country", "year")
  ) %>%
  
  filter(year < 2018) %>%
  na.omit() %>%
  mutate(year = as.numeric(year),
         co2 = as.numeric(co2),
         cdmt = as.numeric(cdmt)) %>%
  mutate(log_co2 = log(co2)) %>%
  mutate(log_cdmt = log(cdmt)) 



```

# Part 2: Linear Regression

## Data Visualizations

### Graph: Relationship Observed Over Time

The first data visualization depicts the relationship between the logged values of Co2 emissions and child mortality rates from 1800 to 2018.

```{r}
#| code-fold: true

co2_cdmt %>%
  group_by(year) %>%
  summarize(avg_co2 = mean(co2),
            avg_cdmt = mean(cdmt)) %>%
  
  ggplot(aes(x = year)) +
  geom_line(aes(y = log(avg_co2), color = "log co2"), size = 1.5) +
  geom_line(aes(y = log(avg_cdmt), color = "log cdmt"), size = 1.5) +labs(title = "Relationship CO2 Emmissions and Child Mortality Rates (log)",
       x = "Year",
       y = "", subtitle = "Emissions in tonnes per capita/proportion of deaths per 1000 births") +
  
  theme_minimal()
```

From this graph we can observe the log transformed data of average carbon dioxide emissions and average child mortality rates from 1800 to 2018 across 192 countries. We logged the data so that we could visually compare the two variables on the same scale. From 1800 to about 1900 is consistently around an average of 6. Just after 1900 the line started to steadily decrease until it almost reached 3 by 2018. The trend for carbon dioxide is slightly different. While there seems to be no change between 1800 and 1825–the emissions are consistently at -3, shortly after that period, the CO2 emissions begin to rapidly increase. There are a few dips in the increasing trend line between 1915 and 1950, but the trend continues to increase. It’s also important to note that there are a few missing variables from the early 2000s. But by 2018 the CO2 emissions have increased to 3. Overtime, as average co2 emissions increase overtime, child mortality rates have decreased.

### Graph: Relationship between CO2 and CDMT

The second visualization is the overall relationship between co2 emissions and child mortality rates by country.

```{r}
#| code-fold: true
graph<-
co2_cdmt %>%
  group_by(country) %>%  
  summarize(log_avg_co2 = mean(log_co2, na.rm = TRUE),
            log_avg_cdmt = mean(log_cdmt, na.rm = TRUE)) 
graph |>
  ggplot(aes(x = log_avg_co2, y = log_avg_cdmt)) +  
  geom_jitter(alpha = 0.7, color = "steelblue", size = 3) +    
  labs(title = "Average CO2 vs. CDMT", 
       subtitle = "Average CDMT", 
       x = "Average CO2",
       y = "") + 
  theme_minimal() +  
  theme(plot.title = element_text(size = 20, face = "bold"),      
        plot.subtitle = element_text(size = 12), 
        axis.text = element_text(size = 10))

```

By observing this graph we can infer an important detail about the relationship between average CO2 emissions and average child mortality rate. Each data point represents a country. It seems as there is a negative relationship between the two variables. As co2 increase child mortality rates decrease.

## Linear Regression

The statistical method we are using to analyze the data is linear regression. This model is used to examine the relationship between an independent and dependent variable.  In this case, CO2 emissions is the independent variable and child mortality rate is the dependent variable. The linear model assumes these two variables have a linear relationship; the changes in the independent variables are associated with a constant change in the dependent variable. This is the form for our model:

y = intercecpt + slope + error term 

Where y represents the dependent variable, child mortality and slope represents in the independent variable, CO2 emissions.

### Graph: Linear Model Visualization

The following graph depicts the visualization of our simple linear regression.

```{r}
#| code-fold: true

graph <- graph %>%
  filter(log_avg_cdmt != -Inf) %>%
  filter(log_avg_co2 != -Inf) 

reg <- lm(log_avg_cdmt~log_avg_co2, data = graph)

graph |>
  ggplot(aes(x=log_avg_co2, y=log_avg_cdmt))+
  geom_jitter()+
  geom_smooth(method = "lm")+ 
  labs(title = "Linear Regression for CO2 and Child Mortality", 
       x = "Average CO2 Emmisions", 
       y = "Average Child Mortality Rate")

```

There appears to be a moderate linear negative relationship between the logged average values of co2 and child mortality. There appears to be outliers on the low end and a slight fanning of the points as co2 increases. There does seem to be a moderate linear correlation between the co2 and cdmt. 

```{r}
#| code-fold: true
broom:: tidy(reg)
```

Estimated Regression Model in Context:

Average Child Mortality Rate = 4.65 - 0.26\*CO2

The intercept is the value of the dependent variable when the independent variable is 0. When CO2 emissions are at 0, the average predicted value for child mortality is 4.65. The slope represents the change in the dependent variable for a one unit increase in the dependent variable. For every one tonne per capita increase in average CO2 emissions the average cdmt decreases by 0.26.

## LINE Conditions

### Linearity and Equal Variance

To assess linearity and equal variance we will asses the residual plot.

```{r}
#| code-fold: true

# Get residuals and predicted values
residual <- augment(reg)
residual_for_norm <- reg$residuals
predicted <- predict(reg)

# Create a residual plot
residual_plot <- ggplot(residual, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Plot",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
print(residual_plot)



```

In this case we observer no residuals above 2 or below -2y o, which reveals no outliers. This plot conveys a slight funnel shape from left to right, that shows more spread on the left and more cluster on the right. Due to this, we cannot state that linearity or equal variance is met. Based on this, we will proceed with caution when fitting a linear model.

### Normality 
```{r}
#| code-fold: true
##qqnorm(residual_for_norm, main = "QQPlot of Residuals")
##qqline(residual_for_norm)

ggplot(data = residual, aes(x = .resid)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Frequency") +
  theme_minimal()
```

Based on the tails of a qqnormal plot we observed, we saw large deviations from the predicted line, meaning that there is a lack of normality. We will proceed with the given data, however have caution when drawing conclusions. We also asses the histogram of the residuals. The distribution of these residuals appears to be bimodal with gaps on either end. Therefore, we conclude the distribution of our residuals are not normal.

### Independence

We do not belive we can assume independence in our data, because Co2 Emission for countries, will likely effect the Co2 emissions of other countries. We think factors like trade and relationships may add confounding variables to our data set.

## Model Fit

```{r}
#| code-fold: true
response <- graph$log_avg_cdmt
fitted_values <- fitted(reg)
residual_2 <- residuals(reg)

var_response <- var(response)
var_fitted <- var(fitted_values)
var_residuals <- var(residual_2)
Poportion_resid <-var_residuals /var_response
variance_table <- data.frame(
  Variable = c("Response", "Fitted Values", "Residuals", "Proportion"),
  Variance = c(var_response, var_fitted, var_residuals, Poportion_resid)
)
print(variance_table)
```

After assessing Linearity,Variance, Normality, and Independence, we do not belive a linear regression model would be appropriate for our data. Additionally, the proportion between the proportion of the variability in the response values that was accounted for by our model, is approximately 0.434. Based on this value we would not assume our data has linear correlation in the slightest. 

# Part 3: Simulation

## Assessing Predictions

### Visualizing Simulations from the Model

```{r}
#| code-fold: true

# Get residuals and predicted values
residual <- augment(reg)
predicted <- predict(reg)

# Calculate sigma of residuals from the regression model
co2_cdmt_sigma <- sigma(reg)

# Define noise function to add random noise
noise <- function(x, mean = 0, sd) {
  x + rnorm(length(x), mean, sd)
}

# Simulate response variable using predicted values and sigma of residuals
sim_response <- tibble(sim_cdmt = noise(predicted, sd = co2_cdmt_sigma))

sim_data <- graph |> 
  select(log_avg_co2, log_avg_cdmt) |> 
  bind_cols(sim_response)


# Plot histograms of observed and simulated response variables
obs_p <- ggplot(data = graph, aes(x = log_avg_co2, y = log_avg_cdmt)) +  
  geom_jitter(alpha = 0.7, color = "steelblue", size = 3) +    
  labs(title = "Observed", 
       subtitle = "Average CDMT", 
       x = "Average CO2",
       y = "")  

new_p <- ggplot(sim_data, aes(x = sim_cdmt, y = log_avg_cdmt )) +
  geom_jitter(alpha = 0.7, color = "steelblue", size = 3) +
  labs(title = "Simulated", y = "", subtitle = " Average CDMT", x = " Average CO2") 

# Combine plots using patchwork
library(patchwork)
obs_p + new_p

```

There is little similarity between the observed and the simulated data leading us to belive, it is not plausible the observed data could be generated by our model. The observed data seems to have a negative correlation, while our simulated data has a positive correlation. This reaffirmed our assumption a linear model is not appropriate to apply to our data.


Adjusted R squared value is 0.555 so there is a very weak linear correlation if that. need more explanation here! HERE: - Write in paragraph form for interpretation of model and description of fit. 

### Inspect R Squared Values

```{r}
#| code-fold: true
nsims <- 1000
sims <- map_dfc(.x = 1:nsims,
                .f = ~ tibble(sim = noise(predicted, 
                                          sd = co2_cdmt_sigma)
                              )
                )

colnames(sims) <- colnames(sims) |> 
  str_replace(pattern = "\\.\\.\\.",
                  replace = "_")

sims <- graph |> 
  filter(!is.na(log_avg_co2), 
         !is.na(log_avg_cdmt)) |> 
  select(log_avg_cdmt) |> 
  bind_cols(sims)


sim_r_sq <- sims |> 
  map(~ lm(log_avg_cdmt ~ .x, data = sims)) |> 
  map(glance) |> 
  map_dbl(~ .x$r.squared)

sim_r_sq <- sim_r_sq[names(sim_r_sq) != "log_avg_cdmt"]

```

```{r}
#| code-fold: true
tibble(sims = sim_r_sq) |> 
  ggplot(aes(x = sims)) + 
  geom_histogram(binwidth = 0.025, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(x = expression("Simulated" ~ R^2),
       y = "Frequency",
       title = "Histogram of Simulated R-squared Values",
       subtitle = "Number of Simulated Models") +
  theme_minimal()
```

This simulated R^2 values seem to be centered around 0.3, with an approximately normal bell shaped curve. This indicates a bad fit, as the values are centered far from 1. Once again, reaffirming our conclusion a linear regression model is not appropriate for our data. In conclusion, C02 Emissions Per Capita is not a good indicator of Child Mortality Rates.
