---
title: "Project Proposal"
author: 'Alexis Hegarty, Ella Smith, Ashlyn Philips, and Amy Truong'
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r}
#| code-fold: true
library(tidyverse)

co2 <- read_csv("co2.csv", col_names = TRUE)
cdmt <- read_csv("cdmt.csv", col_names = TRUE)
```
Hey all, I'm here. It's amy!

hiiii its ash!

### Data Description:

The data we are using in our analysis is sourced from [Gapminder](https://www.gapminder.org/data/documentation/). The Gapminder data sets are compiled of world data ranging from 1800 to 2020. The variables, documented by country (using current geographic boundaries), are indicators of global development and evolution through the past two centuries. Examples of these variables include, child mortality, GDP, fertility rates, life expectancy rates, and more.

The data sets that we will be using in our analysis include two quantitative variables: Carbon Dioxide (CO2) and Child Mortality.

The Child Mortality Rate data set combines data from multiple sources to create a data set of child deaths under the age of 5 for each country from 1800 to 2100. Child mortality is measured as the number of 0-5 years dying per 1000 children born. The data set has three variables: country, year, and child mortality rate.

The Carbon Dioxide data set combines data from multiple sources to create a data set of CO2 emissions in tonnes per capita for each country from 1800 to 2022. The data set has three variables: country, year, and CO2 emissions per capita. Both data sets we are using contain from 194 countries.

### Hypothesis:

We will use CO2 as our explantory variable and Child mortality as our response varaible.

We think that there maybe a relationship between these two variables. We understand that in general CO2 emissions are harmful to humans. From 1800 to 2022 as the world became more industrialized and CO2 emissions per capita exponentially, this also increased child mortality.

However, we are also aware that industrialization also could have lowered child mortality rates through the increase in medicine and technology.

Our investigation aims to explore this nuanced relationship between CO2 emissions and child mortality. Using historical data and statistical analyses, we aim to provide a more comprehensive understanding of the evolution of the relationship between these two variables within the last two decades.

### Cleaning the Data

A glaring factor within the data is the child mortality dataset contains data all the way to 2100. This means the data past 2018 (noted in data description in Gapminder) is predicted. For the purposes of this analysis and the data we have for CO2 emissions, we will only use data from 1800 to 2018.

In our data cleaning process we will omit observations with at least one variable missing. We are aware that omitting these missing variables decreases the sample size and may lead us to miss important information. Since we are using ggplot and lm--which drop missing observations anyway--it makes sense to remove these variables for the entire analysis.

The two data sets will also need to be merged into one data set in long format using the pivot_longer function.

The following code performs these three tasks:

### Merge & Clean Data:

```{r}
#| code-fold: true

co2_cdmt <- cdmt %>%
  mutate(across(-country, as.character)) %>%
  pivot_longer(cols = -country, names_to = 'year', values_to = 'cdmt') %>%
  mutate(year = as.character(year)) %>%
  
  full_join(
    co2 %>%
      mutate(across(-country, as.character)) %>%
      pivot_longer(cols = -country, names_to = 'year', values_to = 'co2') %>%
      mutate(year = as.character(year)),
    by = c("country", "year")
  ) %>%
  
  filter(year < 2018) %>%
  na.omit()


```

### Data Visualization

```{r}
#| code-fold: true

co2_cdmt %>%
  mutate(year = as.numeric(year),
         co2 = as.numeric(co2),
         cdmt = as.numeric(cdmt)) %>%
  group_by(year) %>%
  summarize(avg_co2 = mean(co2),
            avg_cdmt = mean(cdmt)) %>%
  
  ggplot(aes(x = year)) +
  geom_line(aes(y = avg_co2, color = "co2"), size = 1.5) +
  geom_line(aes(y = avg_cdmt, color = "cdmt"), size = 1.5) +labs(title = "Relationship CO2 Emmissions and Child Mortality Rates",
       x = "Year",
       y = "", subtitle = "Emissions in tonnes per capita/proportion of deaths per 1000 births") +
  
  theme_minimal()
```


```{r}
#| code-fold: true

co2_cdmt %>%
  mutate(year = as.numeric(year),
         co2 = as.numeric(co2),
         cdmt = as.numeric(cdmt)) %>%
  group_by(year) %>%
  summarize(avg_co2 = mean(co2),
            avg_cdmt = mean(cdmt)) %>%
  
  ggplot(aes(x = year)) +
  geom_line(aes(y = log(avg_co2), color = "log co2"), size = 1.5) +
  geom_line(aes(y = log(avg_cdmt), color = "log cdmt"), size = 1.5) +labs(title = "Relationship CO2 Emmissions and Child Mortality Rates",
       x = "Year",
       y = "", subtitle = "Emissions in tonnes per capita/proportion of deaths per 1000 births") +
  
  theme_minimal()
```